Magnetised Polish doughnuts revisited

### 2017-07-15 paper_lang_rev/paper_lang_rev.tex
work on language corrections & typo

\sg{Investigation} of \sg{such} systems, either by analytical or numerical means, may rely on the ability to construct suitable 
representations based on physical assumptions. The construction of equilibrium models of stationary discs around black holes has indeed a long tradition (see~\cite{Abramowicz:2013} and references therein). In particular, the \sg{`Polish doughnuts'}~\citep{Abramowicz:1978,Kozlowski:1978} provide a  very general method to build equilibrium configurations of perfect fluid matter orbiting around a Kerr black hole. This fully relativistic model assumes that the disc is non-magnetised and that the matter obeys a  constant specific angular momentum distribution. This method was later extended by~\cite{Komissarov:2006}  by adding a purely azimuthal magnetic field to build magnetised tori around rotating black holes. Dynamical evolutions of magnetized tori built with the Komissarov solution were first reported by~\cite{Montero:2007}.
On the other hand, assuming different distributions of angular momentum in the discs, \citet{Qian:2009} presented a method to build sequences of black hole thick accretion discs in dynamical equilibria, restricted however to the purely hydrodynamical case.

@@

The organization of the paper is as follows: \sg{Sect.}~\ref{framework} presents the analytic framework to build the discs while \sg{Sect.}~\ref{methodology} explains the corresponding numerical procedure. The sequences of models are discussed in \sg{Sect.}~\ref{results}. Finally, the conclusions are summarized in \sg{Sect.}~\ref{conclusions}, where we also briefly indicate potential extensions of this work we plan to inspect. In our work we assume that the test-fluid approximation holds, thus neglecting the self-gravity of the fluid, and further assume that the spacetime is described by the Kerr metric. In mathematical expressions below Greek indices are spacetime indices running from 0 to 3 and Latin indices are only spatial. We use geometrized units where $G = c = 1$.

@@

\begin{table}
\caption{\sg{Models} for $\beta_{\mathrm{m}_{\mathrm{c}}} = 1$. The naming of the models and the parameters reported in the columns are as in Table~\ref{table:1}.}             
\label{table:2}      
\centering          
\begin{tabular}{c c c c  c c c c}
\hline\hline       
 & $\beta$ & $\gamma$ & $r_{\rm{max}}$ &  $r_{\mathrm{m, max}}$ & $\Delta W$               & $r_{\mathrm{in}}$ & $r_{\mathrm{out}}$ \\ 
 &              &                   &                          &                                        & $(\times 10^{-2})$     &                              &  \\
\hline           
A1 & $0.0$ & $0.0$ & $6.00$ &  $6.49$  & $-6.35$ & $2.91$ & -- \\ 
A2 & $0.5$ & $0.5$ & $6.25$ &  $6.66$  & $-2.27$ & $3.20$ & $11.8$\\ 
A3 & $0.9$ & $0.9$ & $6.50$ &  $6.80$  & $-0.30$ & $3.70$ &  $9.58$\\ 
A4 & $0.0$ & $0.9$ & $6.00$ &  $6.49$  & $-6.35$ & $2.91$ & $9.68$\\ 
A5 & $0.9$ & $0.0$ & $6.50$ &  $6.80$  & $-0.30$ & $3.70$ & --\\ 
 \hline 
B1 & $0.0$ & $0.0$ & $3.02$ &  $3.26$  & $-12.9$ & $1.73$ & -- \\ 
B2 & $0.5$ & $0.5$ & $3.16$ &  $3.35$  & $-4.32$ & $1.86$ & $5.35$\\
B3 & $0.9$ & $0.9$ & $3.29$ &  $3.43$  & $-0.54$ & $2.08$ & $4.52$\\ 
B4 & $0.0$ & $0.9$ & $3.02$ &  $3.26$  & $-12.9$ & $1.73$ & $4.54$\\  
B5 & $0.9$ & $0.0$ & $3.29$ &  $3.43$  & $-0.54$ & $2.08$ & -- \\ 
 \hline 
C1 & $0.0$ & $0.0$ & $1.68$ &  $1.80$  & $-0.246$ & $1.21$ & -- \\ 
C2 & $0.5$ & $0.5$ & $1.77$ &  $1.86$  & $-7.34$ & $1.26$ & $2.50$\\  
C3 & $0.9$ & $0.9$ & $1.84$ &  $1.91$  & $-0.85$ & $1.35$ & $2.25$\\ 
C4 & $0.0$ & $0.9$ & $1.68$ &  $1.80$  & $-24.6$ & $1.21$ & $2.26$\\ 
C5 & $0.9$ & $0.0$ & $1.84$ &  $1.91$  & $-0.85$ & $1.35$ & --\\ 
\hline      
\end{tabular}
\end{table}

@@

A representative sample of the isodensity surfaces for some of our models appears in \sg{Fig.}~\ref{models}. We plot, in particular, the models of Table~\ref{table:3}, for which the magnetisation is highest ($\beta_{\mathrm{m}_{\mathrm{c}}}=10^{-3}$). From left to right the columns of this figure correspond to increasing values of the black hole spin, namely, $a=0.5$, 0.9 and 0.99, while from top to bottom the rows correspond to different combinations of the $\gamma$ and $\beta$ parameters that characterize the ansatz for the angular momentum distribution of~\citet{Qian:2009} (the particular values are indicated in the caption of the figure).  A rapid inspection shows that the structure of the discs noticeable changes when the parameters change. Notice that the spatial scale in all of the plots of this figure has been chosen so as to facilitate the visualization of the discs (which can be fairly small in some cases) and, as such, is different in each plot. A typical Polish doughnut is represented by the model in the top-left panel. As the black hole spin increases, the discs are closer to the black hole and its relative size with respect to the black hole is smaller for all values of $\gamma$ and $\beta$. For $\gamma=0$ the discs are infinite, irrespective of the value of $a$ and $\beta$. As $\gamma$ and $\beta$ increase from 0 to 0.9 (compare the three top rows) the discs become significantly thinner and radially smaller (see also Fig.~\ref{more-profile} below). It is also interesting to note that the maximum value of the rest-mass density is higher (with respect to the value at the disc centre) as the spin increases.

@@

Figure~\ref{magnetisation} shows the effects of changing the parameter $\beta_{\mathrm{m}_{\mathrm{c}}}$ (the magnetisation) in the structure of the discs. From left to right the values plotted in each panel are $\beta_{\mathrm{m}_{\mathrm{c}}}=10^3$, 1, and $10^{-3}$. The model chosen corresponds to $a=0.9$ and $\gamma=\beta=0.5$. The larger the value of $\beta_{\mathrm{m}_{\mathrm{c}}}$ the less important the effects of the magnetisation in the disc structure. \sg{Fig.}~\ref{magnetisation} shows that, at least for the kind of magnetic field distribution we are considering, the effects of the magnetisation are minor. The disc structure remains fairly similar for all values of $\beta_{\mathrm{m}_{\mathrm{c}}}
$ and the only quantitative differences are found in the location of the centre of the disc (which moves inward with decreasing $\beta_{\mathrm{m}_{\mathrm{c}}}$) and of the range of variation of the isodensity contours (the maximum being slightly larger with decreasing $\beta_{\mathrm{m}_{\mathrm{c}}}$). This can be more clearly visualized in \sg{Fig.}~\ref{magnetisation-profile} which displays the radial profile at the equatorial plane of the rest-mass density for the same three cases plotted in Fig.~\ref{magnetisation}.

>>>>

\sg{Investigation} of \sg{such} systems, either by analytical or numerical means, may rely on the ability to construct suitable 
representations based on physical assumptions. The construction of equilibrium models of stationary discs around black holes has indeed a long tradition (see~\cite{Abramowicz:2013} and references therein). In particular, the \sg{`Polish doughnuts'}~\citep{Abramowicz:1978,Kozlowski:1978} provide a  very general method to build equilibrium configurations of perfect fluid matter orbiting around a Kerr black hole. This fully relativistic model assumes that the disc is non-magnetised and that the matter obeys a  constant specific angular momentum distribution. This method was later extended by~\cite{Komissarov:2006}  by adding a purely azimuthal magnetic field to build magnetised tori around rotating black holes. Dynamical evolutions of \sg{magnetised} tori built with the Komissarov solution were first reported by~\cite{Montero:2007}.
On the other hand, assuming different distributions of angular momentum in the discs, \citet{Qian:2009} presented a method to build sequences of black hole thick accretion discs in dynamical equilibria, restricted however to the purely hydrodynamical case.

@@

The organization of the paper is as follows: \sg{Sect.}~\ref{framework} presents the analytic framework to build the discs while \sg{Sect.}~\ref{methodology} explains the corresponding numerical procedure. The sequences of models are discussed in \sg{Sect.}~\ref{results}. Finally, the conclusions are \sg{summarised} in \sg{Sect.}~\ref{conclusions}, where we also briefly indicate potential extensions of this work we plan to inspect. In our work we assume that the test-fluid approximation holds, thus neglecting the self-gravity of the fluid, and further assume that the spacetime is described by the Kerr metric. In mathematical expressions below Greek indices are spacetime indices running from 0 to 3 and Latin indices are only spatial. We use \sg{geometrised} units where $G = c = 1$.

@@

\begin{table}
\caption{\sg{Models} for $\beta_{\mathrm{m}_{\mathrm{c}}} = 1$. The naming of the models and the parameters reported in the columns are as in Table~\ref{table:1}.}             
\label{table:2}      
\centering          
\begin{tabular}{c c c c  c c c c}
\hline\hline       
 & $\beta$ & $\gamma$ & $r_{\rm{max}}$ &  $r_{\mathrm{m, max}}$ & $\Delta W$               & $r_{\mathrm{in}}$ & $r_{\mathrm{out}}$ \\ 
 &              &                   &                          &                                        & $(\times 10^{-2})$     &                              &  \\
\hline           
A1 & $0.0$ & $0.0$ & $6.00$ &  $6.49$  & $-6.35$ & $2.91$ & -- \\ 
A2 & $0.5$ & $0.5$ & $6.25$ &  $6.66$  & $-2.27$ & $3.20$ & $11.8$\\ 
A3 & $0.9$ & $0.9$ & $6.50$ &  $6.80$  & $-0.30$ & $3.70$ &  $9.58$\\ 
A4 & $0.0$ & $0.9$ & $6.00$ &  $6.49$  & $-6.35$ & $2.91$ & $9.68$\\ 
A5 & $0.9$ & $0.0$ & $6.50$ &  $6.80$  & $-0.30$ & $3.70$ & --\\ 
 \hline 
B1 & $0.0$ & $0.0$ & $3.02$ &  $3.26$  & $-12.9$ & $1.73$ & -- \\ 
B2 & $0.5$ & $0.5$ & $3.16$ &  $3.35$  & $-4.32$ & $1.86$ & $5.35$\\
B3 & $0.9$ & $0.9$ & $3.29$ &  $3.43$  & $-0.54$ & $2.08$ & $4.52$\\ 
B4 & $0.0$ & $0.9$ & $3.02$ &  $3.26$  & $-12.9$ & $1.73$ & $4.54$\\  
B5 & $0.9$ & $0.0$ & $3.29$ &  $3.43$  & $-0.54$ & $2.08$ & -- \\ 
 \hline 
C1 & $0.0$ & $0.0$ & $1.68$ &  $1.80$  & $-24.6$ & $1.21$ & -- \\ 
C2 & $0.5$ & $0.5$ & $1.77$ &  $1.86$  & $-7.34$ & $1.26$ & $2.50$\\  
C3 & $0.9$ & $0.9$ & $1.84$ &  $1.91$  & $-0.85$ & $1.35$ & $2.25$\\ 
C4 & $0.0$ & $0.9$ & $1.68$ &  $1.80$  & $-24.6$ & $1.21$ & $2.26$\\ 
C5 & $0.9$ & $0.0$ & $1.84$ &  $1.91$  & $-0.85$ & $1.35$ & --\\ 
\hline      
\end{tabular}
\end{table}

@@

A representative sample of the isodensity surfaces for some of our models appears in \sg{Fig.}~\ref{models}. We plot, in particular, the models of Table~\ref{table:3}, for which the magnetisation is highest ($\beta_{\mathrm{m}_{\mathrm{c}}}=10^{-3}$). From left to right the columns of this figure correspond to increasing values of the black hole spin, namely, $a=0.5$, 0.9 and 0.99, while from top to bottom the rows correspond to different combinations of the $\gamma$ and $\beta$ parameters that \sg{characterise} the ansatz for the angular momentum distribution of~\citet{Qian:2009} (the particular values are indicated in the caption of the figure).  A rapid inspection shows that the structure of the discs noticeable changes when the parameters change. Notice that the spatial scale in all of the plots of this figure has been chosen so as to facilitate the visualization of the discs (which can be fairly small in some cases) and, as such, is different in each plot. A typical Polish doughnut is represented by the model in the top-left panel. As the black hole spin increases, the discs are closer to the black hole and its relative size with respect to the black hole is smaller for all values of $\gamma$ and $\beta$. For $\gamma=0$ the discs are infinite, irrespective of the value of $a$ and $\beta$. As $\gamma$ and $\beta$ increase from 0 to 0.9 (compare the three top rows) the discs become significantly thinner and radially smaller (see also Fig.~\ref{more-profile} below). It is also interesting to note that the maximum value of the rest-mass density is higher (with respect to the value at the disc centre) as the spin increases.

@@

Figure~\ref{magnetisation} shows the effects of changing the parameter $\beta_{\mathrm{m}_{\mathrm{c}}}$ (the magnetisation) in the structure of the discs. From left to right the values plotted in each panel are $\beta_{\mathrm{m}_{\mathrm{c}}}=10^3$, 1, and $10^{-3}$. The model chosen corresponds to $a=0.9$ and $\gamma=\beta=0.5$. The larger the value of $\beta_{\mathrm{m}_{\mathrm{c}}}$ the less important the effects of the magnetisation in the disc structure. \sg{Fig.}~\ref{magnetisation} shows that, at least for the kind of magnetic field distribution we are considering, the effects of the magnetisation are minor. The disc structure remains fairly similar for all values of $\beta_{\mathrm{m}_{\mathrm{c}}}
$ and the only quantitative differences are found in the location of the centre of the disc (which moves inward with decreasing $\beta_{\mathrm{m}_{\mathrm{c}}}$) and of the range of variation of the isodensity contours (the maximum being slightly larger with decreasing $\beta_{\mathrm{m}_{\mathrm{c}}}$). This can be more clearly \sg{visualised} in \sg{Fig.}~\ref{magnetisation-profile} which displays the radial profile at the equatorial plane of the rest-mass density for the same three cases plotted in Fig.~\ref{magnetisation}.
